Software Architectures to Integrate Workflow Engines in Science Gateways

### 2016-04-28 paper.tex
Renamed 'sub-tasking' with 'task encapsulation', following @merous' comment.

\definecolor{ao(english)}{rgb}{0.0, 0.5, 0.0}
\newcommand{\todo}[2]{\pdfmargincomment[color=red,author=#1,open=true]{#2}}
\newcommand{\note}[2]{\pdfmargincomment[color=yellow,author=#1,open=true]{#2}}
\newcommand{\closednote}[4]{\pdfmargincomment[color=gray,author=#1-#3,open=false]{#1: #2 \textCR \textCR #3: #4}}
\newcommand{\answerednote}[4]{\pdfmargincomment[color=green,author=#1-#3,open=true]{#1: #2 \textCR \textCR #3: #4}}
\newcommand{\closedanswerednote}[6]{\pdfmargincomment[color=gray,author=#1-#3-#5,open=true]{#1: #2 \textCR \textCR #3: #4 \textCR \textCR #5: #6}}

@@

\begin{abstract}
  We investigate software architectures commonly used to integrate
  workflow engines in science gateways. We abstract software
  components and interactions in 6 architectures that we use to
  classify 18 systems. We compare these architectures using metrics
  measuring global complexity, robustness, extensibility, scalability
  and other specific features. Results show the pros and cons of the
  different architectures: tight integration and sub-tasking are the
  simplest architectures and should therefore be the easiest to
  implement and the most robust ones, most architectures perform the
  same in terms of extensibility, the pool model is the most scalable
  and meta-workflows are only available in nested workflows and
  workflow import. Our analysis provides insights for software
  architects on how to integrate science gateways and workflow
  engines.
\end{abstract}

@@

\begin{figure*}
\centering
\hspace*{0.2\columnwidth}
\subfloat[Tight integration]{
\def\svgwidth{0.6\columnwidth}
\input{figures/tight.pdf_tex}
\label{archi:tight}
}
\hfill
\subfloat[Service invocation]{
\def\svgwidth{0.6\columnwidth}
\input{figures/service.pdf_tex}
\label{archi:service}
\hspace*{0.2\columnwidth}
}\\
\hspace*{0.2\columnwidth}
\subfloat[Sub-tasking]{
\def\svgwidth{0.6\columnwidth}
\input{figures/sub-task.pdf_tex}
\label{archi:sub-task}
}
\hfill
\subfloat[Pool]{
\def\svgwidth{0.6\columnwidth}
\input{figures/agent.pdf_tex}
\label{archi:agent}
\hspace*{0.2\columnwidth}
}\\
\subfloat[Nested workflows. Left: abstract model. Right: instantiation with service invocation.]{
\def\svgwidth{0.9\columnwidth}
\input{figures/nested-1.pdf_tex}
\def\svgwidth{0.9\columnwidth}
\input{figures/nested-2.pdf_tex}
\label{archi:nested}
}\\
\subfloat[Workflow import. Left: abstract model. Right: instantiation with service invocation.]{
\def\svgwidth{0.5\columnwidth}
\input{figures/import-1.pdf_tex}
\hfill
\def\svgwidth{0.5\columnwidth}
\input{figures/import-2.pdf_tex}
\label{archi:import}
}
\caption{Architectures}
\label{fig:architectures}
\end{figure*}
\begin{table*}
\centering
\footnotesize
\rowcolors{1}{white}{gray!25}
\begin{tabular}{ll}
  \textbf{Architecture} & \textbf{Systems} \\
  \hline
  Tight & \pbox{1.5\columnwidth}{
          Catania Science Gateway Framework~\cite{Ardizzone2012},
          Distributed application runtime environment (DARE~\cite{maddineni2012distributed}),
          DECIDE~\cite{ardizzone2012decide}$^{c,n}$, LONI Pipeline Environment$^n$~\cite{dinov2009efficient}
          }
  \\
  Service & \pbox{1.5\columnwidth}{
            Apache Airvata~\cite{marru2011apache}, e-BioInfra~\cite{shahand2015data}$^{w,n}$, HubZero with Pegasus~\cite{CPE:CPE3257}, MoSGrid~\cite{krüger2014mosgrid}$^w$, System in~\cite{wu2010accelerating}, Vine Toolkit~\cite{DBLP:journals/scpe/SzejnfeldDKKKKLPTWDNW10}, Virtual Imaging Platform~\cite{GLAT-13}$^n$, WS-PGRADE/gUSE framework~\cite{Kacsuk2012},
            Science gateways in~\cite{kacsuk2014science}$^w$
            } \\
  Sub-task & CBRAIN and PSOM~\cite{GLAT-16}$^n$, CBRAIN and FSL$^n$\\
  Pool & SHIWA pool~\cite{ROGE-13}\\
  Nested & \pbox{1.5\columnwidth}{SHIWA Simulation Platform (Coarse-Grained Interoperability~\cite{terstyanszky2014enabling})$^w$, HubZero with Pegasus (via hierarchical workflows)~\cite{Deelman201517}, Tavaxy~\cite{Abouelhoda2012}.
           }\\
  Import & SHIWA Simulation Platform (Fine-Grained Interoperability)~\cite{plankensteiner-prodan-etal:2013}$^w$, Tavaxy~\cite{Abouelhoda2012}, system in~\cite{delaGarza2016}.
\end{tabular}
\caption{Classification of science gateways based on the architecture used to integrate workflow engines. $^c$: based on the Catania Science Gateway Framework. $^w$: based on WS-PGRADE/gUSE. $^n$: used for neuroimaging.}
\label{table:system-classification}
\end{table*}

@@

\subsection{Sub-tasking} \note{Marc-e}{May I propose another name here? Either
"Task-centric" or "Task Oriented". My reasoning is that the WE is just another
task, as well as the sub-tasks themselves.}

See Figure~\subref*{archi:sub-task}. The workflow engine is a particular
task that can submit sub-tasks to the science gateway through
interaction \texttt{e}. The workflow engine keeps track of the
dependencies between the sub-tasks, but their execution is delegated to
the science gateway that executes them on the infrastructure through
interaction \texttt{b}. Although the science gateway has no global
vision of the workflow, it can keep track of the sub-tasks submitted
by a given task, for instance to be able to cancel them when the task
is canceled. The science gateway may also implement mechanisms to
facilitate the handling of task dependencies, for instance basic
dependency lists as available in most batch managers (see for instance
attribute \texttt{depend} of option \texttt{-W} in
Torque\footnote{\url{http://docs.adaptivecomputing.com}}).

@@

The sub-tasking architecture is implemented in CBRAIN~\cite{SHER-14}
where it is used to integrate the FSL toolkit~\cite{Jenkinson2012782}
and the PSOM workflow engine~\cite{bellec2012pipeline}.

@@

The CBRAIN-PSOM integration~\cite{GLAT-16} is shown on
Figure~\ref{fig:cbrain-psom-architecture}. The PSOM workflow engine
adopts a pilot-job architecture~\cite{turilli2015comprehensive} where
a master coordinates workflow execution by submitting workers and
establishing direct communication channels with them. Note how this
peculiar execution model is well supported by the sub-tasking
architecture.

\begin{figure*}
\centering
\subfloat[CBRAIN-FSL integration.]{
\def\svgwidth{\columnwidth}
\input{figures/CBRAIN-FSL.pdf_tex}
\label{fig:cbrain-fsl-architecture}
} \hfill \subfloat[CBRAIN-PSOM integration. Figure
reproduced from~\cite{GLAT-16}]{ \def\svgwidth{\columnwidth}
  \input{figures/CBRAIN-PSOM.pdf_tex}
\label{fig:cbrain-psom-architecture}
}
\caption{CBRAIN architecture for workflow engine integration
  (sub-tasking).  1: User sends data and workflow execution request to
  storage server(s) and CBRAIN portal. 2: CBRAIN portal sends
  execution request to execution server on cluster. 3: Execution
  server transfers data from storage server(s). 4-5: Execution server
  starts workflow engine (FSL tool or PSOM master) via task
  scheduler. 6: Workflow engine submits sub-tasks to execution server
  (FSL tasks or PSOM agents). 7-8: Execution server starts sub-tasks
  through task scheduler. FSL sub-tasks will run locally
  instead of being submitted to CBRAIN through interaction 6. 8':
  (PSOM only) PSOM master and workers execute workflow. 9. Execution
  server transfers results to storage server(s). Interaction
  \texttt{b} in Figure~\ref{fig:architectures} is implemented by steps
  \texttt{4}, \texttt{5}, \texttt{7} and \texttt{8} (regular
  interactions with batch manager). Interaction \texttt{c} is
  implemented by steps \texttt{3} and \texttt{9}. Interaction
  \texttt{e} is implemented by step \texttt{6} (for FSL: through a
  modified version of the fsl\_sub script available in
  \url{https://github.com/aces/cbrain-plugins-neuro}; for PSOM:
  through a specific development in PSOM).}
\label{fig:cbrain-sub-tasking}
\end{figure*}

@@

                                     & \textbf{Sub-task}

@@

\subsection{Sub-task}

\paragraph{Global complexity} The sub-task architecture requires only 2
components (\texttt{I$_1$}=2).  It involves 5 interactions:
\texttt{a$_1$}, \texttt{a$_2$}, \texttt{b}, \texttt{c} and
\texttt{e} (\texttt{I$_2$}=5).

@@

\paragraph{Scalability}
New engine instances are spawned and executed on the infrastructure as
any other task upon user submission (\texttt{S$_1$}=0). This is a
major interest of the sub-task architecture. Distributed engines are
not supported by default (\texttt{S$_2$}=1). Task scheduling is
slightly more complex than in the other approaches due to the special
role of the task that executes the workflow engine
(\texttt{S$_3$}=1). Indeed, the reliability of this task is critical
since all the sub-tasks in the workflow depend on it and, depending on
the recovery capabilities of the workflow engine, may need to be
resubmitted if the workflow task fails. The workflow task is also
longer than all its sub-tasks, which increases its chances of
failure. In addition, task parameters, for instance estimated
walltime, are more difficult to estimate for the workflow task than
for sub-tasks which may generate issues such as selection of wrong
batch queues on clusters. Finally, the interdependencies between the
workflow task and its sub-tasks may create deadlocks when there is
contention. For instance, if only 1 computing resource is
available for the science gateway and that the workflow task is
running on it and submits sub-tasks, then the sub-tasks could only
execute when the resource is available, which will never happen
because the workflow task will not complete until the sub-tasks
complete. This configuration can be generalized to an infrastructure
with $n$ resources where $n$ workflows are submitted. In practice,
however, the number of submitted workflows usually remains lower than
the number of computing resources available on this infrastructure,
which makes such deadlocks unlikely to happen.

\subsection{Pool}

@@

Tight integration and sub-tasking are globally the simplest
architectures, followed by service integration, workflow import (with
service invocation) and pool. Nested workflows (with service
invocation) requires more integration than the other
architectures. Robustness roughly leads to the same ordering of
architectures, with tight integration and sub-tasking in the top
group, service integration and workflow import (with service
invocation) close behind, pool in a third group, and nested workflows
(with service invocation) at the end. This ordering is consistent
across metrics, it reflects the global complexity of the
architectures.

Regarding extensibility, most architectures are overall comparable,
except nested workflows (with service invocation) which is
significantly behind. This is explained by the complexity of the
nested workflow architecture, with 2 infrastructures and 2 workflow
services. Sub-tasking and workflow import (with service invocation)
perform a bit better when integrating new engines (\texttt{E$_1$}),
which is interesting given the current profusion of engines with
different characteristics.  However, sub-tasking, workflow import
(with service invocation) and tight integration perform worse than the
others when adding engine versions (\texttt{E$_2$}) due to the need to
update infrastructure (sub-tasking), science gateway (tight
integration), or workflow converter (workflow import). Workflow import
performs worse than the others when integrating new workflows
(\texttt{E$_3$}) because of the language conversion step. All
architectures except nested workflows perform the same when
integrating new infrastructures (\texttt{E$_4$}).

The pool architecture is overall the most scalable, which is not a
surprise since it was designed precisely for scalability. Tight
integration is the least scalable and all the other architectures
perform the same. The overall scalability score, however, should not
mask the unique characteristics of architectures regarding this
criterion. Nested workflows are the only architecture that can easily
accommodate distributed workflow execution, which can be critical in
some cases. At the same time, the scheduling constraints created by
the sub-tasking and nested workflow architectures could well become
showstoppers depending on the type of the targeted
infrastructure. Non-reliable infrastructures, for example, could
hardly cope with workflow engines being wrapped in computing tasks as
done in sub-tasking. The availability of multiple engine instances
could also become a critical feature for science gateways with
important workloads, which would favor sub-tasking and pool.

Differences in other specific features should not be neglected. Nested
workflows (with service invocation) and workflow import (with service
invocation) are the only architectures that support meta-workflows,
which may be interesting in some cases. Also, sub-tasking and nested
workflows may complexify fine-grained debugging. \closednote{Rafael}{I would
  suggest to provide some more examples when this could be
  interesting.}{Tristan}{Done.}  The availability of fine-grained debugging
information may also be critical to efficient user support.

@@

favor some of them. Overall, sub-tasking, and workflow import (with

@@

A few limitations remain that should be considered when using the
results of our evaluation. First of all, the evaluation is done at a
quite high-level of abstraction that may be a bit artificial when
applied to real systems. Abstracting architectures from their
implementation in specific systems is useful to compare them, but we
also realize that this comes at the cost of realism. In particular,
the presented architectures are theoritical patterns that may be blended in
actual systems. The distinction between tight integration and service
invocation, for instance, may not be always that clear in practice. Service
invocation may also be combined with sub-tasking in some cases. However, the
criteria discussed in this work would still apply to broadly compare and categorize
such cases of hydrid architectures.

@@

The abstract nested workflows and workflow import patterns were
instantiated with service invocation so that they can be analyzed in
the same framework as the other architectures. Other types of
instantiation, for instance nested workflows with sub-tasking, could
also be envisaged. We chose to limit ourselves to instantiations with
service invocation because the resulting architectures are implemented
in real systems, and because service invocation is largely
used. Nested workflows with sub-tasking, in particular, could be an
interesting architecture to explore.

The particular technical or historical context of a science gateway
project may obviously influence the choice of an architecture to
integrate workflow engines. For instance, some workflow engines may be
already available as web services, which would tend to favor service
invocation, and other science gateways may have strongly tested task
and data control features (interactions \texttt{b} and \texttt{c}),
which would favor sub-tasking. Similarly, adding a new type of engine
may facilitate integration of a new infrastructure when interactions
\texttt{b} and \texttt{c$_2$} are already available. The migration
cost between architectures has been ignored as well.

>>>>

\definecolor{ao(english)}{rgb}{0.0, 0.5, 0.0}
\newcommand{\todo}[2]{\pdfmargincomment[color=red,author=#1,open=true]{#2}}
\newcommand{\note}[2]{\pdfmargincomment[color=yellow,author=#1,open=true]{#2}}
\newcommand{\closednote}[4]{} % closed notes are hidden. This should obviously be a parameter of \note rather than a separate command.
\newcommand{\answerednote}[4]{\pdfmargincomment[color=green,author=#1-#3,open=true]{#1: #2 \textCR \textCR #3: #4}}
\newcommand{\closedanswerednote}[6]{\pdfmargincomment[color=gray,author=#1-#3-#5,open=true]{#1: #2 \textCR \textCR #3: #4 \textCR \textCR #5: #6}}

@@

\begin{abstract}
  We investigate software architectures commonly used to integrate
  workflow engines in science gateways. We abstract software
  components and interactions in 6 architectures that we use to
  classify 18 systems. We compare these architectures using metrics
  measuring global complexity, robustness, extensibility, scalability
  and other specific features. Results show the pros and cons of the
  different architectures: tight integration and task encapsulation are the
  simplest architectures and should therefore be the easiest to
  implement and the most robust ones, most architectures perform the
  same in terms of extensibility, the pool model is the most scalable
  and meta-workflows are only available in nested workflows and
  workflow import. Our analysis provides insights for software
  architects on how to integrate science gateways and workflow
  engines.
\end{abstract}

@@

\begin{figure*}
\centering
\hspace*{0.2\columnwidth}
\subfloat[Tight integration]{
\def\svgwidth{0.6\columnwidth}
\input{figures/tight.pdf_tex}
\label{archi:tight}
}
\hfill
\subfloat[Service invocation]{
\def\svgwidth{0.6\columnwidth}
\input{figures/service.pdf_tex}
\label{archi:service}
\hspace*{0.2\columnwidth}
}\\
\hspace*{0.2\columnwidth}
\subfloat[Task encapsulation]{
\def\svgwidth{0.6\columnwidth}
\input{figures/sub-task.pdf_tex}
\label{archi:sub-task}
}
\hfill
\subfloat[Pool]{
\def\svgwidth{0.6\columnwidth}
\input{figures/agent.pdf_tex}
\label{archi:agent}
\hspace*{0.2\columnwidth}
}\\
\subfloat[Nested workflows. Left: abstract model. Right: instantiation with service invocation.]{
\def\svgwidth{0.9\columnwidth}
\input{figures/nested-1.pdf_tex}
\def\svgwidth{0.9\columnwidth}
\input{figures/nested-2.pdf_tex}
\label{archi:nested}
}\\
\subfloat[Workflow import. Left: abstract model. Right: instantiation with service invocation.]{
\def\svgwidth{0.5\columnwidth}
\input{figures/import-1.pdf_tex}
\hfill
\def\svgwidth{0.5\columnwidth}
\input{figures/import-2.pdf_tex}
\label{archi:import}
}
\caption{Architectures}
\label{fig:architectures}
\end{figure*}
\begin{table*}
\centering
\footnotesize
\rowcolors{1}{white}{gray!25}
\begin{tabular}{ll}
  \textbf{Architecture} & \textbf{Systems} \\
  \hline
  Tight & \pbox{1.5\columnwidth}{
          Catania Science Gateway Framework~\cite{Ardizzone2012},
          Distributed application runtime environment (DARE~\cite{maddineni2012distributed}),
          DECIDE~\cite{ardizzone2012decide}$^{c,n}$, LONI Pipeline Environment$^n$~\cite{dinov2009efficient}
          }
  \\
  Service & \pbox{1.5\columnwidth}{
            Apache Airvata~\cite{marru2011apache}, e-BioInfra~\cite{shahand2015data}$^{w,n}$, HubZero with Pegasus~\cite{CPE:CPE3257}, MoSGrid~\cite{krüger2014mosgrid}$^w$, System in~\cite{wu2010accelerating}, Vine Toolkit~\cite{DBLP:journals/scpe/SzejnfeldDKKKKLPTWDNW10}, Virtual Imaging Platform~\cite{GLAT-13}$^n$, WS-PGRADE/gUSE framework~\cite{Kacsuk2012},
            Science gateways in~\cite{kacsuk2014science}$^w$
            } \\
  Task & CBRAIN and PSOM~\cite{GLAT-16}$^n$, CBRAIN and FSL$^n$\\
  Pool & SHIWA pool~\cite{ROGE-13}\\
  Nested & \pbox{1.5\columnwidth}{SHIWA Simulation Platform (Coarse-Grained Interoperability~\cite{terstyanszky2014enabling})$^w$, HubZero with Pegasus (via hierarchical workflows)~\cite{Deelman201517}, Tavaxy~\cite{Abouelhoda2012}.
           }\\
  Import & SHIWA Simulation Platform (Fine-Grained Interoperability)~\cite{plankensteiner-prodan-etal:2013}$^w$, Tavaxy~\cite{Abouelhoda2012}, system in~\cite{delaGarza2016}.
\end{tabular}
\caption{Classification of science gateways based on the architecture used to integrate workflow engines. $^c$: based on the Catania Science Gateway Framework. $^w$: based on WS-PGRADE/gUSE. $^n$: used for neuroimaging.}
\label{table:system-classification}
\end{table*}

@@

\subsection{Task encapsulation} \answerednote{Marc-e}{May I propose
  another name here? Either "Task-centric" or "Task Oriented". My
  reasoning is that the WE is just another task, as well as the
  sub-tasks themselves.}{Tristan}{I agree: sub-tasking is not the
  integration model, it is the process that engines use when they
  are integrated as tasks. I think 'task encapsulation' is more
  consistent  with 'service invocation' than 'task-oriented'}

See Figure~\subref*{archi:sub-task}. The workflow engine is wrapped in
a particular task that can submit sub-tasks to the science gateway
through interaction \texttt{e}. The workflow engine keeps track of the
dependencies between the sub-tasks, but their execution is delegated
to the science gateway that executes them on the infrastructure
through interaction \texttt{b}. Although the science gateway has no
global vision of the workflow, it can keep track of the sub-tasks
submitted by a given task, for instance to be able to cancel them when
the task is canceled. The science gateway may also implement
mechanisms to facilitate the handling of task dependencies, for
instance basic dependency lists as available in most batch managers
(see for instance attribute \texttt{depend} of option \texttt{-W} in
Torque\footnote{\url{http://docs.adaptivecomputing.com}}).

@@

Task encapsulation is implemented in CBRAIN~\cite{SHER-14} where it is
used to integrate the FSL toolkit~\cite{Jenkinson2012782} and the PSOM
workflow engine~\cite{bellec2012pipeline}.

@@

The CBRAIN-PSOM integration~\cite{GLAT-16} is shown on
Figure~\ref{fig:cbrain-psom-architecture}. The PSOM workflow engine
adopts a pilot-job architecture~\cite{turilli2015comprehensive} where
a master coordinates workflow execution by submitting workers and
establishing direct communication channels with them. Note how this
peculiar execution model is well supported by task encapsulation.

\begin{figure*}
\centering
\subfloat[CBRAIN-FSL integration.]{
\def\svgwidth{\columnwidth}
\input{figures/CBRAIN-FSL.pdf_tex}
\label{fig:cbrain-fsl-architecture}
} \hfill \subfloat[CBRAIN-PSOM integration. Figure
reproduced from~\cite{GLAT-16}]{ \def\svgwidth{\columnwidth}
  \input{figures/CBRAIN-PSOM.pdf_tex}
\label{fig:cbrain-psom-architecture}
}
\caption{CBRAIN architecture for workflow engine integration
  (task encapsulation).  1: User sends data and workflow execution request to
  storage server(s) and CBRAIN portal. 2: CBRAIN portal sends
  execution request to execution server on cluster. 3: Execution
  server transfers data from storage server(s). 4-5: Execution server
  starts workflow engine (FSL tool or PSOM master) via task
  scheduler. 6: Workflow engine submits sub-tasks to execution server
  (FSL tasks or PSOM agents). 7-8: Execution server starts sub-tasks
  through task scheduler. FSL sub-tasks will run locally
  instead of being submitted to CBRAIN through interaction 6. 8':
  (PSOM only) PSOM master and workers execute workflow. 9. Execution
  server transfers results to storage server(s). Interaction
  \texttt{b} in Figure~\ref{fig:architectures} is implemented by steps
  \texttt{4}, \texttt{5}, \texttt{7} and \texttt{8} (regular
  interactions with batch manager). Interaction \texttt{c} is
  implemented by steps \texttt{3} and \texttt{9}. Interaction
  \texttt{e} is implemented by step \texttt{6} (for FSL: through a
  modified version of the fsl\_sub script available in
  \url{https://github.com/aces/cbrain-plugins-neuro}; for PSOM:
  through a specific development in PSOM).}
\label{fig:cbrain-sub-tasking}
\end{figure*}

@@

                                     & \textbf{Task}

@@

\subsection{Task encapsulation}

\paragraph{Global complexity} Task encapsulation requires only 2
components (\texttt{I$_1$}=2).  It involves 5 interactions:
\texttt{a$_1$}, \texttt{a$_2$}, \texttt{b}, \texttt{c} and
\texttt{e} (\texttt{I$_2$}=5).

@@

\paragraph{Scalability}
New engine instances are spawned and executed on the infrastructure as
any other task upon user submission (\texttt{S$_1$}=0). This is a
major interest of task encapsulation. Distributed engines are
not supported by default (\texttt{S$_2$}=1). Task scheduling is
slightly more complex than in the other approaches due to the special
role of the task that executes the workflow engine
(\texttt{S$_3$}=1). Indeed, the reliability of this task is critical
since all the sub-tasks in the workflow depend on it and, depending on
the recovery capabilities of the workflow engine, may need to be
resubmitted if the workflow task fails. The workflow task is also
longer than all its sub-tasks, which increases its chances of
failure. In addition, task parameters, for instance estimated
walltime, are more difficult to estimate for the workflow task than
for sub-tasks which may generate issues such as selection of wrong
batch queues on clusters. Finally, the interdependencies between the
workflow task and its sub-tasks may create deadlocks when there is
contention. For instance, if only 1 computing resource is
available for the science gateway and that the workflow task is
running on it and submits sub-tasks, then the sub-tasks could only
execute when the resource is available, which will never happen
because the workflow task will not complete until the sub-tasks
complete. This configuration can be generalized to an infrastructure
with $n$ resources where $n$ workflows are submitted. In practice,
however, the number of submitted workflows usually remains lower than
the number of computing resources available on this infrastructure,
which makes such deadlocks unlikely to happen.

\subsection{Pool model}

@@

Tight integration and task encapsulation are globally the simplest
architectures, followed by service integration, workflow import (with
service invocation) and pool. Nested workflows (with service
invocation) requires more integration than the other
architectures. Robustness roughly leads to the same ordering of
architectures, with tight integration and task encapsulation in the top
group, service integration and workflow import (with service
invocation) close behind, pool in a third group, and nested workflows
(with service invocation) at the end. This ordering is consistent
across metrics, it reflects the global complexity of the
architectures.

Regarding extensibility, most architectures are overall comparable,
except nested workflows (with service invocation) which is
significantly behind. This is explained by the complexity of the
nested workflow architecture, with 2 infrastructures and 2 workflow
services. Task encapsulation and workflow import (with service invocation)
perform a bit better when integrating new engines (\texttt{E$_1$}),
which is interesting given the current profusion of engines with
different characteristics.  However, task encapsulation, workflow import
(with service invocation) and tight integration perform worse than the
others when adding engine versions (\texttt{E$_2$}) due to the need to
update infrastructure (task encapsulation), science gateway (tight
integration), or workflow converter (workflow import). Workflow import
performs worse than the others when integrating new workflows
(\texttt{E$_3$}) because of the language conversion step. All
architectures except nested workflows perform the same when
integrating new infrastructures (\texttt{E$_4$}).

The pool architecture is overall the most scalable, which is not a
surprise since it was designed precisely for scalability. Tight
integration is the least scalable and all the other architectures
perform the same. The overall scalability score, however, should not
mask the unique characteristics of architectures regarding this
criterion. Nested workflows are the only architecture that can easily
accommodate distributed workflow execution, which can be critical in
some cases. At the same time, the scheduling constraints created by
task encapsulation and nested workflows could well become
showstoppers depending on the type of the targeted
infrastructure. Non-reliable infrastructures, for example, could
hardly cope with workflow engines being wrapped in computing tasks as
done in task encapsulation. The availability of multiple engine instances
could also become a critical feature for science gateways with
important workloads, which would favor task encapsulation and pool.

Differences in other specific features should not be neglected. Nested
workflows (with service invocation) and workflow import (with service
invocation) are the only architectures that support meta-workflows,
which may be interesting in some cases. Also, task encapsulation and nested
workflows may complexify fine-grained debugging. \closednote{Rafael}{I would
  suggest to provide some more examples when this could be
  interesting.}{Tristan}{Done.}  The availability of fine-grained debugging
information may also be critical to efficient user support.

@@

favor some of them. Overall, task encapsulation, and workflow import (with

@@

A few limitations remain that should be considered when using the
results of our evaluation. First of all, the evaluation is done at a
quite high-level of abstraction that may be a bit artificial when
applied to real systems. Abstracting architectures from their
implementation in specific systems is useful to compare them, but we
also realize that this comes at the cost of realism. In particular,
the presented architectures are theoritical patterns that may be blended in
actual systems. The distinction between tight integration and service
invocation, for instance, may not be always that clear in practice. Service
invocation may also be combined with task encapsulation in some cases. However, the
criteria discussed in this work would still apply to broadly compare and categorize
such cases of hydrid architectures.

@@

The abstract nested workflows and workflow import patterns were
instantiated with service invocation so that they can be analyzed in
the same framework as the other architectures. Other types of
instantiation, for instance nested workflows with task encapsulation, could
also be envisaged. We chose to limit ourselves to instantiations with
service invocation because the resulting architectures are implemented
in real systems, and because service invocation is largely
used. Nested workflows with task encapsulation, in particular, could be an
interesting architecture to explore.

The particular technical or historical context of a science gateway
project may obviously influence the choice of an architecture to
integrate workflow engines. For instance, some workflow engines may be
already available as web services, which would tend to favor service
invocation, and other science gateways may have strongly tested task
and data control features (interactions \texttt{b} and \texttt{c}),
which would favor task encapsulation. Similarly, adding a new type of engine
may facilitate integration of a new infrastructure when interactions
\texttt{b} and \texttt{c$_2$} are already available. The migration
cost between architectures has been ignored as well.
